---
title: "Regression and other stories"
subtitle: "Andrew Gelman, Jennifer Hill, Aki Vehtari"
author: "Dan Killian notes - Chapter 1"
date: "August 2020"
output: 
  html_document:
    toc: true
    toc_depth: '5'
    toc_float: yes
---

## 

```{r global_options, echo=F}

# standard figure size and generate clean output
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align='center', warning=FALSE, message=FALSE, cache=TRUE, error=T)

```

  
```{r libraries, echo=F}

library(tidyverse)
library(strengejacke)
library(psych)
library(scales)
library(brms)
library(rethinking)
library(rstanarm)
library(rstantools)
library(skimr)
library(knitr)
library(xlsx)
library(here)
library(haven)
library(showtext)
library(gfonts)

```

```{r formatting, echo=F}

# display to three decimal points, scientific notation displays at around six digits
options(digits=3, scipen=6)

# requires showtext package
showtext_auto()

# requires gfonts package
use_pkg_gfont("open-sans")

# graphic style
theme_set(theme_bw() + theme(panel.grid.minor.x=element_blank(),
                             panel.grid.minor.y=element_blank(),
                             plot.title=element_text(face="bold",size=18, hjust=.5, family = "sans"),
                             plot.subtitle = element_text(size=16, family="sans"),
                             plot.caption=element_text(size=12, family="sans"),
                             axis.title=element_text(size=16, family="sans"),
                             axis.text=element_text(size=14, family="sans"),    
                             legend.text=element_text(size=14, family="sans"),
                             strip.text=element_text(size=14, family="sans"))) 

# for single graphs, I have recently come to like borderless
nobord <- theme(panel.border=element_blank(),
                axis.ticks = element_blank())


```

# Chapter 1 - Prediction as a unifying theme in statistics and causal inference
## 1.1 The three challenges of statistics

- Generalizing from sample to population

- Generalizing from treatment to control group

- Generalizing from observed measurements to the underlying constructs of interest

Ky skills needed to address these challenges: 

- Understanding regression models

- Constructing regression models

- Fitting regression models to data

- Displaying and interpreting results


Figure 1.1 generated in section 1.2

## 1.2 Why learn regression?

Regression is a method that allows researchers to summarize how predictions or average values of an outcome vary across individuals defined by a set of predictors. 

```{r}
hibbs <- read.table("ElectionsEconomy/data/hibbs.dat", header=T)
head(hibbs)
```

```{r, results="hide"}
growth <- stan_glm(vote ~ growth, 
                   data=hibbs)
```

```{r}
summary(growth)
```

```{r, fig.height=6, fig.width=8, fig.align='center'}
ggplot(hibbs, aes(growth/100, vote/100, label=year)) +
  geom_hline(yintercept=.5, color="darkgoldenrod3", alpha=.4, linetype="dotdash", size=1.3) + 
  stat_smooth(color="darkblue", alpha=.6, se=F, method="lm") + 
  geom_label(size=3, color="darkblue") +
  nobord + 
  scale_x_continuous(labels=percent_format(accuracy=1)) +
  scale_y_continuous(labels=percent_format(accuracy=1)) + 
  labs(x="Growth rate",
         y="Incumbent vote share",
         title="Economic growth and incumbent vote share") + 
  annotate("text", x=.01, y=.58, label="incumbent vote share = 46.3 + 3 * growth rate")
```

The uses of regression: 

- Prediction

- Exploring associations

- Extrapolation

- Causal inference

## 1.3 Some uses of regression

### Figure 1.2

```{r}
elec_wide <- read.table("ElectricCompany/data/electric_wide.txt", header=TRUE)
head(elec_wide)
```

```{r}
out <- elec_wide %>%
  select(2, 4, 6) %>%
  pivot_longer(cols=2:3,
               names_to="round")
```

```{r}
head(out)
```

Get means for grade and observational status. 

```{r}
mns <- out %>%
  group_by(grade, round) %>%
  summarize(score=mean(value))
mns
```

```{r fig.width=12}
ggplot(out, aes(x=value)) + 
  geom_histogram(fill="midnight blue", color="cyan", alpha=.7, bins=40, binwidth=2) + 
  facet_grid(round ~ grade) +
  geom_vline(aes(xintercept=score), data=mns, size=1.2, color="maroon") +
  scale_x_continuous(breaks=c(50,75,100)) +
  theme_bw() + 
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_text(angle=0, vjust=.5, hjust=1),
        strip.background.y=element_blank(),
        strip.text.y=element_blank()) +
  labs(x="",
       y="Control\nclasses\n\n\n\n\n\n\n\n\n\n\n\nTreatment\nclasses",
       title="Post test scores, by grade and observational status",
       caption="Electric company evaluation")
```


### Figure 1.3


```{r}
peace <- read_dta("Peacekeeping/data/pk&pkept_old.dta")
```

```{r, eval=T}
peace2 <- peace %>%
  mutate(delay = (as.numeric(faildate) - as.numeric(cfdate)) /365.24,
         badness=(log(hazard1) / 2) + 8,
         trt=factor(pk_dum, labels=c("No peacekeeping","Peacekeeping"))) %>%
    filter(morewar==1,
           pcw==1,
           !is.na(delay)) %>%
  set_labels(pk_dum, labels=c("No peacekeeping","Peacekeeping"))
                      
```


```{r}

ggplot(peace2, aes(x=delay)) + 
  geom_histogram(fill="midnight blue", color="cyan", alpha=.7, bins=20, binwidth=.5) + 
  facet_wrap(~trt) +
    theme_bw() + 
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_text(angle=0, vjust=.5, hjust=1),
        strip.background.y=element_blank(),
        strip.text.y=element_blank()) +
  labs(y="",
       x="Years until return of war",
       title="Return to war after no peacekeeping, peacekeeping",
       caption="Countries without return to war:\n56% with peacekeeping\n36% without peacekeeping")

```

### Figure 1.4

```{r}
ggplot(peace2, aes(badness, delay)) + 
  geom_point() + 
  stat_smooth(method="lm", se=F) + 
  facet_wrap(~trt) +
  labs(x="Index of pre-treatment characteristics\n(badness)",
       y="Years before return to war",
       title="Effect of peacekeeping on reversion to civil war") + 
  theme_bw()
```

## 1.4 Challenges in building, understanding, and interpreting regressions

### Figure 1.5

```{r}
set.seed(5324)
N <- 50
dat <- data.frame(id=1:50) %>%
  mutate(x=runif(N,1,5),
         y=rnorm(N, 10+3*x, 3),
         xbin = ifelse(x<3, 0,1),
         y2 = rnorm(5 + 30*exp(-x), 2))
head(dat)
```

```{r, results="hide"}
d1 <- stan_glm(y ~ xbin, 
               data=dat)
```

```{r}
summary(d1)
```


```{r}
ggplot(dat, aes(xbin, y)) + 
  geom_point(size=1, color="darkblue", alpha=.4) + 
  stat_smooth(method="lm", se=F, color="darkblue") +
  theme_bw() +
  nobord +
  scale_x_continuous(limits=c(-.1,1.1),
                     breaks=c(0,1),
                     labels=c("Comparison","Treatment")) + 
  scale_y_continuous(breaks=seq(0,30,5),
                     limits=c(0,30)) +
  labs(x="",
       y="Outcome",
       title="Regression with binary treatment") + 
  annotate("text", x=.6, y=7.5, label="Estimated treatment effect\nis slope of fitted line: 6.0")
```

```{r, results="hide"}
d2 <- stan_glm(y ~ x,
               data=dat)
a <- tidy(d2)
```

```{r}
summary(d2)
```

```{r}
ggplot(dat, aes(x, y)) + 
  geom_point(size=1, color="darkblue", alpha=.4) + 
  stat_smooth(method="lm", se=F, color="darkblue") +
  theme_bw() +
  nobord +
  scale_x_continuous(limits=c(1,5),
                     breaks=1:5) + 
  scale_y_continuous(breaks=seq(0,30,5),
                     limits=c(0,30)) +
  labs(x="",
       y="Outcome",
       title="Regression with continuous treatment") 
```

### Figure 1.6

```{r, results="hide"}
d3 <- stan_glm(y2 ~ x,
               data=dat)
```

```{r}
summary(d3)
```

```{r}
ggplot(dat, aes(x, y2)) + 
  geom_point(size=1, color="darkblue", alpha=.4) + 
  stat_smooth(se=F, color="darkblue") +
  theme_bw() +
  nobord +
  scale_x_continuous(limits=c(1,5),
                     breaks=1:5) + 
  scale_y_continuous(breaks=seq(0,5,1),
                     limits=c(0,5)) +
  labs(x="",
       y="Outcome",
       title="Nonlinear treatment effect")
```

```{r}
ggplot(dat, aes(x, y2)) + 
  geom_point(size=1, color="darkblue", alpha=.4) + 
  stat_smooth(method="lm", se=F, color="darkblue") +
  theme_bw() + 
  nobord +
  scale_x_continuous(limits=c(1,5),
                     breaks=1:5) + 
  scale_y_continuous(breaks=seq(0,5,1),
                     limits=c(0,5)) +
  labs(x="",
       y="Outcome",
       title="Nonlinear effect, estimated with straight line") 
```

### Figure 1.7

### Figure 1.8

```{r}
set.seed(432)
N <- 100
xx <- rnorm(N, 0, 1)^2
z <- rep(0:1, N/2)
xx <- ifelse(z==0, rnorm(N, 0, 1.2)^2, rnorm(N, 0, .8)^2)
yy <- rnorm(N, 20 + 5*xx + 10*z, 3)
d3 <- data.frame(xx, z, yy) %>%
  mutate(z=factor(z,
                  labels=c("Comparison","Treatment")))
head(d3)
```

```{r, results="hide"}
d <- stan_glm(yy ~ xx + z,
              data=d3)
```

```{r}
summary(d)
```


```{r}
ggplot(d3, aes(xx, yy, color=z)) + 
  #geom_point(alpha=.6, size=1) + 
  stat_smooth(method="lm", se=F) +
  #stat_smooth(aes(yy), data=filter(d3, z==1)) +
  #stat_smooth(aes(yy), data=filter(d3, z==0)) +
  scale_x_continuous(breaks=0:12) +
  scale_color_manual(values=c("maroon","darkblue")) +
  theme_bw() +
  theme(legend.title=element_blank(),
        legend.position="bottom") +
  labs(x="Multi-valued pre-treatment predictor",
       y="Outcome",
       title="Treatment effect with pre-treatment predictor") +
  annotate("text", x=8.1, y=35, label="Treatment effect: 10") +
  nobord
```















